---
title: ''
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load required packages
library(jsonlite)
library(tidyverse)
library(keyring)
library(janitor)

#store your census api key - only needs to be done once per computer
#key_set("census-api-key")

```

## Pull table B01001 (sex by age) from 5 year ACS for all CCDPH municipalities

```{r acs_age_sex_muni}

#load in ccdph muni file with place code
munis <- read_csv("https://github.com/Cook-County-Department-of-Public-Health/ccdph-data-sets/blob/main/Municipality%20Populations.csv?raw=TRUE") %>% 
  #filter(ExcludeFromAnalysis == F) %>%   
  mutate(CensusPlaceCode = ifelse(nchar(CensusPlaceCode) == 4, paste0("0", CensusPlaceCode), CensusPlaceCode))

#load in ACS variable names to match with codes
acs_vars <- fromJSON("https://api.census.gov/data/2019/acs/acs5/variables") %>%
  as.data.frame() %>%
  row_to_names(row_number = 1)

#Create empty list to store place data
acs_age_pops_muni_list <- list()

#loop through munis list to pull data
for (i in 1:nrow(munis)) {
  
  #determine if partial place string needed
  partial <- munis[[i, "Partial"]]
  
  #extract place code  -- NOTE: ifelse should not necessary if performed when importing the muni file, remove after next test
  #place_code <- as.character(munis[[i, "CensusPlaceCode"]] %>% ifelse(nchar(.) == 4, paste0("0", .), .))
  place_code <- munis[[i, "CensusPlaceCode"]]
  
  #construct api call - different string needed for if pulling for only part of town within cook county
  if (partial) {
    api <- paste0("https://api.census.gov/data/2019/acs/acs5?get=NAME,group(B01001)&for=county%20(or%20part):031&in=state:17%20place:", place_code, "&key=", key_get("census-api-key"))
  } else {
    api <- paste0("https://api.census.gov/data/2019/acs/acs5?get=NAME,group(B01001)&for=place:", place_code, "&in=state:17&key=", key_get("census-api-key"))
  }
  
  #pull data
  town_pull <- fromJSON(api) %>% 
    as.data.frame() %>%
    row_to_names(row_number = 1) %>%
    pivot_longer(cols = -NAME, names_to = "variable", values_to = "estimate") %>%
    filter(grepl("*(E|M)$", variable)) %>%    #keep only population estimates and MOEs (see note below, no annonations present)
    separate(variable, into = c("variable", "estimate_type"), sep = -1) %>%
    pivot_wider(c(NAME, variable), names_from = estimate_type, values_from = estimate) %>%
    rename(estimate = E, margin_of_error = M) %>%
    mutate(variable = paste0(variable, "E")) %>%
    left_join(acs_vars, by = c("variable" = "name")) %>%   #link back to acs list of variables for label to ensure expected variables are pulled
    mutate(CensusPlaceCode = as.character(place_code))
  
  #add results to list
  acs_age_pops_muni_list[[i]] <- town_pull
  
}

#combine separate town files into one
acs_age_pops_muni <- bind_rows(acs_age_pops_muni_list)

#first exploration - see if any estimates annotated -- No annotations present, don't need to include in final loop
#acs_age_pops_muni %>% filter(grepl("M", variable) & !is.na(estimate)) %>% View()

#QUALITY CHECK -- make sure sum of pops add up to the whole
check_pops <- acs_age_pops_muni %>%
  filter(!variable %in% c("B01001_001E", "B01001_002E", "B01001_026E"))  %>% #remove total pop, total male, total female 
  group_by(NAME) %>%
  summarize(total_pop_calc = sum(as.numeric(estimate))) %>%  #sum all mutually exclusive male and female age groups for calculated total pop
  ungroup() %>%
  left_join(filter(acs_age_pops_muni, variable == "B01001_001E")) %>%
  mutate(diff = total_pop_calc - as.numeric(estimate))  #verify all diff values equal zero

#QULAITY CHECK -- make sure place code matching expected name from muni file 
name_compare <- acs_age_pops_muni %>%
  filter(variable == "B01001_001E") %>%
  left_join(munis, by = "CensusPlaceCode") %>%
  mutate(clean_census_name = gsub(" village, Illinois| city, Illinois| town, Illinois", "", NAME),
         clean_census_name = gsub("^.*, ", "", clean_census_name),
         fips_check = trimws(clean_census_name) == trimws(Municipality)) %>%
  select(NAME, CensusPlaceCode, clean_census_name, Municipality, fips_check)

#QULAITY CHECK -- compare ACS total pops to decennial total pops to check for any wild discrepancies
total_pop_compare <- acs_age_pops_muni %>%
  filter(variable == "B01001_001E") %>%
  mutate(estimate = as.numeric(estimate)) %>%
  left_join(munis, by = "CensusPlaceCode") %>%
  mutate(diff = estimate - Population2010,
         percent_change = ((estimate - Population2010) / Population2010) * 100,
         percent_change = round(percent_change, 2),
         percent_change_abs = abs(percent_change)) %>%
  select(NAME, Municipality, CensusPlaceCode, Population2010, estimate, diff, percent_change, percent_change_abs) 


#Store data set in github repository
#Format determined from consensus discussion in inter-epi group
acs_age_pop_muni_clean <- acs_age_pops_muni %>%
  filter(!variable %in% c("B01001_001E", "B01001_002E", "B01001_026E"))  %>% #remove totals
  mutate(label = gsub("Estimate!!Total:!!", "", label)) %>%
  separate(label, into = c("sex", "age"), sep = ":!!") %>%
  mutate(clean_name = gsub(" village, Illinois| city, Illinois| town, Illinois", "", NAME),
         clean_name = gsub("^.*, ", "", clean_name)) %>%
  select(census_name = NAME,
         clean_name,
         census_place_code = CensusPlaceCode, 
         variable,
         sex,
         age,
         estimate,
         margin_of_error)

write_csv(acs_age_pop_muni_clean, "acs-5yr-2019-age-sex-by-muni.csv")

#Explore margins of error
explore_moe <- acs_age_pop_muni_clean %>% 
  mutate(estimate = as.numeric(as.character(estimate)), 
         margin_of_error = as.numeric(as.character(margin_of_error)), 
         moe_bigger = margin_of_error > estimate, 
         moe_percent = (margin_of_error / estimate) * 100)

explore_moe %>% filter(moe_bigger) %>% nrow() #16% of cells have MOEs larger than the estimate
explore %>% filter(moe_percent > 50) %>% nrow() #45% of cells have MOEs larger than half the estimate


```

## Pull table P12 (sex by age) from decennial 2010 summary file 1 for all CCDPH municipalities

*Note: API call for partial Homer Glen fails - not sure why. Call works for 2019 5 YR ACS (0 people within Cook County). Maybe decennial doesn't have any part of it in Cook County?  Filter out for now.*

```{r sf1_age_sex_muni}

#load in ccdph muni file with place code
munis <- read_csv("https://github.com/Cook-County-Department-of-Public-Health/ccdph-data-sets/blob/main/Municipality%20Populations.csv?raw=TRUE") %>% 
  #filter(ExcludeFromAnalysis == F) %>%   
  mutate(CensusPlaceCode = ifelse(nchar(CensusPlaceCode) == 4, paste0("0", CensusPlaceCode), CensusPlaceCode)) %>%
  filter(Municipality != "Homer Glen")  #see Note above

#load in sf1 variable names to match with codes
dec_vars <- fromJSON("https://api.census.gov/data/2010/dec/sf1/variables") %>%
  as.data.frame() %>%
  row_to_names(row_number = 1)

#Create empty list to store place data
dec_age_pops_muni_list <- list()

#loop through munis list to pull data
for (i in 1:nrow(munis)) {
  
  #determine if partial place string needed
  partial <- munis[[i, "Partial"]]
  
  #extract place code  -- NOTE: ifelse should not necessary if performed when importing the muni file, remove after next test
  #place_code <- as.character(munis[[i, "CensusPlaceCode"]] %>% ifelse(nchar(.) == 4, paste0("0", .), .))
  place_code <- munis[[i, "CensusPlaceCode"]]
  
  #construct api call - different string needed for if pulling for only part of town within cook county
  if (partial) {
    api <- paste0("https://api.census.gov/data/2010/dec/sf1?get=NAME,group(P12)&for=county%20(or%20part):031&in=state:17%20place:", place_code, "&key=", key_get("census-api-key"))
  } else {
    api <- paste0("https://api.census.gov/data/2010/dec/sf1?get=NAME,group(P12)&for=place:", place_code, "&in=state:17&key=", key_get("census-api-key"))
  }
  
  #pull data
  town_pull <- fromJSON(api) %>% 
    as.data.frame() %>%
    row_to_names(row_number = 1) %>%
    pivot_longer(cols = -NAME, names_to = "variable", values_to = "estimate") %>%
    filter(grepl("^P.*", variable)) %>%    #keep only population estimates, no meta data
    filter(!grepl(".*ERR$", variable)) %>%
    left_join(dec_vars, by = c("variable" = "name")) %>%   #link back to acs list of variables for label to ensure expected variables are pulled
    mutate(CensusPlaceCode = as.character(place_code)) %>%
    mutate(estimate = as.numeric(as.character(estimate)))  #numbers being stored as factors
  
  #add results to list
  dec_age_pops_muni_list[[i]] <- town_pull
  
}

#combine separate town files into one
dec_age_pops_muni <- bind_rows(dec_age_pops_muni_list)

#QUALITY CHECK -- make sure sum of pops add up to the whole
check_pops <- dec_age_pops_muni %>%
  filter(!variable %in% c("P012001", "P012002", "P012026"))  %>% #remove total pop, total male, total female 
  group_by(NAME) %>%
  summarize(total_pop_calc = sum(estimate)) %>%  #sum all mutually exclusive male and female age groups for calculated total pop
  ungroup() %>%
  left_join(filter(dec_age_pops_muni, variable == "P012001")) %>%
  mutate(diff = total_pop_calc - as.numeric(estimate))  #verify all diff values equal zero

#QULAITY CHECK -- make sure place code matching expected name from muni file 
name_compare <- dec_age_pops_muni %>%
  filter(variable == "P012001") %>%
  left_join(munis, by = "CensusPlaceCode") %>%
  mutate(clean_census_name = gsub(" village, Illinois| city, Illinois| town, Illinois", "", NAME),
         clean_census_name = gsub("^.*, ", "", clean_census_name),
         fips_check = trimws(clean_census_name) == trimws(Municipality)) %>%
  select(NAME, CensusPlaceCode, clean_census_name, Municipality, fips_check)


#NOTE: when pulling 2020 data, may want to adapt this quality check to compare 2020 decennial to 2010 decennial for wild discrepancies
total_pop_compare <- dec_age_pops_muni %>%
  filter(variable == "P012001") %>%
  left_join(munis, by = "CensusPlaceCode") %>%
  mutate(diff = estimate - Population2010) %>%
  select(NAME, Municipality, CensusPlaceCode, Population2010, estimate, diff) 


#Store data set in github repository
#Format determined from consensus discussion in inter-epi group
dec_age_pops_muni_clean <- dec_age_pops_muni %>%
  filter(!variable %in% c("P012001", "P012002", "P012026"))  %>% #remove total pop, total male, total female 
  mutate(label = gsub("Total!!", "", label)) %>%
  separate(label, into = c("sex", "age"), sep = "!!") %>%
  select(census_name = NAME,
         census_place_code = CensusPlaceCode, 
         variable,
         sex,
         age,
         estimate)

write_csv(dec_age_pops_muni_clean, "decennial-2010-age-sex-by-muni.csv")

```


## Pull table P5 (Hispanic or Latino Origin by Race) from decennial 2010 summary file 1 for all CCDPH municipalities


```{r sf1_race_muni}

#load in ccdph muni file with place code
munis <- read_csv("https://github.com/Cook-County-Department-of-Public-Health/ccdph-data-sets/blob/main/Municipality%20Populations.csv?raw=TRUE") %>% 
  #filter(ExcludeFromAnalysis == F) %>%   
  mutate(CensusPlaceCode = ifelse(nchar(CensusPlaceCode) == 4, paste0("0", CensusPlaceCode), CensusPlaceCode)) %>%
  filter(Municipality != "Homer Glen")  #see Note above age

#load in sf1 variable names to match with codes
dec_vars <- fromJSON("https://api.census.gov/data/2010/dec/sf1/variables") %>%
  as.data.frame() %>%
  row_to_names(row_number = 1)

#Create empty list to store place data
dec_race_pops_muni_list <- list()

#loop through munis list to pull data
for (i in 1:nrow(munis)) {
  
  #determine if partial place string needed
  partial <- munis[[i, "Partial"]]
  
  #extract place code  -- NOTE: ifelse should not necessary if performed when importing the muni file, remove after next test
  #place_code <- as.character(munis[[i, "CensusPlaceCode"]] %>% ifelse(nchar(.) == 4, paste0("0", .), .))
  place_code <- munis[[i, "CensusPlaceCode"]]
  
  #construct api call - different string needed for if pulling for only part of town within cook county
  if (partial) {
    api <- paste0("https://api.census.gov/data/2010/dec/sf1?get=NAME,group(P5)&for=county%20(or%20part):031&in=state:17%20place:", place_code, "&key=", key_get("census-api-key"))
  } else {
    api <- paste0("https://api.census.gov/data/2010/dec/sf1?get=NAME,group(P5)&for=place:", place_code, "&in=state:17&key=", key_get("census-api-key"))
  }
  
  #pull data
  town_pull <- fromJSON(api) %>% 
    as.data.frame() %>%
    row_to_names(row_number = 1) %>%
    pivot_longer(cols = -NAME, names_to = "variable", values_to = "estimate") %>%
    filter(grepl("^P.*", variable)) %>%    #keep only population estimates, no meta data
    filter(!grepl(".*ERR$", variable)) %>%
    left_join(dec_vars, by = c("variable" = "name")) %>%   #link back to acs list of variables for label to ensure expected variables are pulled
    mutate(CensusPlaceCode = as.character(place_code)) %>%
    mutate(estimate = as.numeric(as.character(estimate)))  #numbers being stored as factors
  
  #add results to list
  dec_race_pops_muni_list[[i]] <- town_pull
  
}

#combine separate town files into one
dec_race_pops_muni <- bind_rows(dec_race_pops_muni_list)

#QUALITY CHECK -- make sure sum of pops add up to the whole
check_pops <- dec_race_pops_muni %>%
  filter(!variable %in% c("P005001", "P005002", "P005010"))  %>% #remove total pop, total non-hispanic, total hispanic 
  group_by(NAME) %>%
  summarize(total_pop_calc = sum(estimate)) %>%  #sum all mutually exclusive race/ethnicity groups for calculated total pop
  ungroup() %>%
  left_join(filter(dec_race_pops_muni, variable == "P005001")) %>%
  mutate(diff = total_pop_calc - estimate)  #verify all diff values equal zero

#QULAITY CHECK -- make sure place code matching expected name from muni file 
name_compare <- dec_race_pops_muni %>%
  filter(variable == "P005001") %>%
  left_join(munis, by = "CensusPlaceCode") %>%
  mutate(clean_census_name = gsub(" village, Illinois| city, Illinois| town, Illinois", "", NAME),
         clean_census_name = gsub("^.*, ", "", clean_census_name),
         fips_check = trimws(clean_census_name) == trimws(Municipality)) %>%
  select(NAME, CensusPlaceCode, clean_census_name, Municipality, fips_check)


#QULAITY CHECK -- compare ACS total pops to decennial total pops to check for any wild discrepancies
#NOTE: when pulling 2020 data, may want to adapt this quality check to compare 2020 decennial to 2010 decennial for wild discrepancies
total_pop_compare <- dec_race_pops_muni %>%
  filter(variable == "P005001") %>%
  left_join(munis, by = "CensusPlaceCode") %>%
  mutate(diff = estimate - Population2010) %>%
  select(NAME, Municipality, CensusPlaceCode, Population2010, estimate, diff) 

#Store data set for future use -- modify storage location
#write_csv(dec_race_pops_muni, "decennial-2010-race-by-muni.csv")
#Store data set in github repository
#Format determined from consensus discussion in inter-epi group
dec_race_pops_muni_clean <- dec_race_pops_muni %>%
  filter(!variable %in% c("P005001", "P005002", "P005010"))  %>% #remove total pop, total non-hispanic, total hispanic  
  mutate(label = gsub("Total!!", "", label)) %>%
  separate(label, into = c("ethnicity", "race"), sep = "!!") %>%
  mutate(clean_name = gsub(" village, Illinois| city, Illinois| town, Illinois", "", NAME),
         clean_name = gsub("^.*, ", "", clean_name)) %>%
  select(census_name = NAME,
         clean_name,
         census_place_code = CensusPlaceCode, 
         variable,
         ethnicity,
         race,
         estimate)

write_csv(dec_race_pops_muni_clean, "decennial-2010-race-ethnicity-by-muni.csv")


```

## Pull total population from decennial 2010 summary file 1 for all ZCTAs that have population in Cook County

```{r sf1_total_zcta}

#import current zip code file
current_zips <- read_csv("https://github.com/Cook-County-Department-of-Public-Health/ccdph-data-sets/blob/main/Suburban%20Cook%20Zip%20Codes.csv?raw=TRUE") %>% 
  clean_names() %>%
  mutate(zip_code = as.character(zip_code))

#list zctas in illinois (note:only one IL ZCTA crosses state lines, down state so not of concern for CCDPH)
zcta_il <- fromJSON(paste0("https://api.census.gov/data/2010/dec/sf1?get=P001001,NAME&for=zip%20code%20tabulation%20area%20(or%20part):*&in=state:17&key=", key_get("census-api-key"))) %>%
  as.data.frame() %>%
  row_to_names(row_number = 1) %>%
  select(zcta = `zip code tabulation area (or part)`,
         population = P001001)

#function to pull pop in cook
get_partial <- function(zcta){
  
  value <- fromJSON(paste0("https://api.census.gov/data/2010/dec/sf1?get=P001001,NAME&for=county%20(or%20part):031&in=state:17%20zip%20code%20tabulation%20area%20(or%20part):", zcta, "&key=", key_get("census-api-key")))[2,1]
  
  return(as.numeric(value))
  
}

#Note: code below works but is very slow (~25 minutes); might want to for loop with try() at some point
zcta_pops <- zcta_il %>%
  rowwise() %>%
  mutate(population_cook = possibly(get_partial, otherwise = NA)(zcta))

zcta_pops_clean <- zcta_pops %>%
  drop_na(population_cook) %>%
  mutate(population = as.numeric(population),
         percent_in_cook = (population_cook / population) * 100,
         percent_in_cook = round(percent_in_cook, 1),
         partial = population != population_cook) %>%
  full_join(current_zips, by = c("zcta" = "zip_code")) %>%
  mutate(primary_city = case_when(!is.na(primary_city) ~ primary_city,
                                  grepl("^606.*", zcta) ~ "Chicago",
                                  grepl("^603.*", zcta) ~ "Oak Park",
                                  grepl("^602.*", zcta) ~ "Evanston",
                                  grepl("^6007.*", zcta) ~ "Skokie",
                                  zcta == "60459" ~ "Burbank")) %>%
  mutate(district = ifelse(is.na(district) & primary_city %in% c("Chicago", "Oak Park", "Evanston", "Skokie", "Stickney", "Burbank", "Forest View"), "OOJ", district)) %>%
  #rename(ccdph_district = district) %>%
  mutate(nssp_region = as.character(NA)) %>%
  mutate(notes = case_when(zcta == "60007" ~ "borders cross into O'Hare and DuPage but all residents in Cook per census",
                           zcta == "60402" ~ "part in Berwyn and part in Stickney Township")) %>%  #add institutional knowledge previously stored in previous zip code file's "cross_borders" field
  select(-c(population_zcta_2010, cross_borders)) %>%
  select(zip_code = zcta,
         zcta_total_pop = population,
         zcta_pop_in_cook = population_cook,
         percent_pop_cook = percent_in_cook,
         partial,
         district,
         primary_city,
         nssp_region,
         notes)
  
#write file to csv and manually fill in missing district and primary city for partial zips using melissa data
write_csv(zcta_pops_clean, "decennial-2010-total-pop-zcta.csv")

```

## Pull single year age pops <20 for CCDPH jurisdiction

```{r sf1_ccdph_single_age}



```


## Pull 12+ population by muni

```{r sf2_twelve_plus}

#load in ccdph muni file with place code
munis <- read_csv("https://github.com/Cook-County-Department-of-Public-Health/ccdph-data-sets/blob/main/Municipality%20Populations.csv?raw=TRUE") %>% 
  #filter(ExcludeFromAnalysis == F) %>%   
  mutate(CensusPlaceCode = ifelse(nchar(CensusPlaceCode) == 4, paste0("0", CensusPlaceCode), CensusPlaceCode)) %>%
  filter(!ExcludeFromAnalysis)

#load in sf2 variable names to match with codes
dec_vars <- fromJSON("https://api.census.gov/data/2010/dec/sf2/variables") %>%
  as.data.frame() %>%
  row_to_names(row_number = 1)


#loop through munis list to pull data
over_12_by_muni = sapply(munis$CensusPlaceCode, function(x){
  print(x) #for debugging and checking progress
  
  #determine if partial place string needed
  partial = munis$Partial[munis$CensusPlaceCode == x]
  
  #extract place code 
  place_code = x
  
  #construct api call - different string needed for if pulling for only part of town within cook county
  if (partial) {
    api <- paste0("https://api.census.gov/data/2010/dec/sf2?get=NAME,group(PCT3)&for=county%20(or%20part):031&in=state:17%20place:", place_code, "&key=", key_get("census-api-key"))
    } else {
    api <- paste0("https://api.census.gov/data/2010/dec/sf2?get=NAME,group(PCT3)&for=place:", ... = place_code, "&in=state:17&key=", key_get("census-api-key"))
    }
  
  #pull data
  town_pull <- fromJSON(api) %>% 
    as.data.frame() %>%
    row_to_names(row_number = 1) %>%
    clean_names(case = "all_caps") %>%
    filter(POPGROUP_TTL == "Total population") %>%
    select(-NAME_2, GEO_ID, POPGROUP, POPGROUP_TTL, STATE, PLACE) %>%
    pivot_longer(cols = -NAME, names_to = "variable", values_to = "estimate") %>%
    filter(grepl("^PCT.*", variable)) %>%    #keep only population estimates, no meta data
    filter(!grepl(".*ERR$", variable)) %>%
    left_join(dec_vars, by = c("variable" = "name")) %>%   #link back to acs list of variables for label to ensure expected variables are pulled
    mutate(CensusPlaceCode = as.character(place_code)) %>%
    mutate(estimate = as.numeric(as.character(estimate))) %>% #numbers being stored as factors
    filter(grepl("\\d{2,}", label),
           !grepl("10 years|11 years", label)
           ) #filter to 12+
  
  over_12_pop = sum(town_pull$estimate)

  return(over_12_pop)
  
})

#combine separate town files into one
over_12_by_muni_table = munis %>%
  left_join(bind_cols("CensusPlaceCode" = names(over_12_by_muni), 
                                  "Population2010_Over12" = as.numeric(over_12_by_muni)),
                      by = "CensusPlaceCode") %>%
  mutate(PercentOver12 = round(Population2010_Over12/Population2010*100, 2)) %>%
  select(Municipality, CensusPlaceCode, Population2010, Population2010_Over12, PercentOver12)

write_csv(over_12_by_muni_table, "decennial-2010-over-12-by-muni.csv")  

#over 12 overall
devtools::source_url("https://github.com/Cook-County-Department-of-Public-Health/ccdph-functions/blob/master/census-functions.R?raw=TRUE")

over_12_vars = dec_vars %>%
  filter(concept == "SEX BY AGE") %>%
  filter(grepl("\\d{2,}", label),
           !grepl("10 years|11 years", label)
           ) %>% #filter to 12+
  rowwise() %>%
  mutate(ccdph_population = ccdph_census_population(name, sumfile = "sf2"))
  

ccdph_over_12 = sum(over_12_vars$ccdph_population) #1909596
```


#Single age pops for ccdph jurisdiction
```{r sf2_single_age_pops}
#load in sf2 variable names
dec_vars <- fromJSON("https://api.census.gov/data/2010/dec/sf2/variables") %>%
  as.data.frame() %>%
  row_to_names(row_number = 1)

devtools::source_url("https://github.com/Cook-County-Department-of-Public-Health/ccdph-functions/blob/master/census-functions.R?raw=TRUE")

#single age pops overall ccdph jurisdiction
single_age_pops = dec_vars %>%
  filter(concept == "SEX BY AGE") %>%
  filter(grepl("\\d", label)) %>% #filter out summaries
  mutate(sex = ifelse(grepl("Female", label), "Female", "Male"),
         age = gsub("^.*ale!!", "", label)
         ) %>%
  rowwise() %>%
  mutate(ccdph_population_2010 = ccdph_census_population(name, sumfile = "sf2")) %>%
  arrange(name)

write_csv(single_age_pops, "decennial-2010-age-sex-single-age-ccdph-jurisdiction.csv")

```



```{r general_examples}

#pull decennial variable for all of cook county
fromJSON(paste0("https://api.census.gov/data/2010/dec/sf1?get=P001001,NAME&for=county:031&in=state:17&key=", key_get("census-api-key")))

#pull decennial variable for place within cook county -- NOTE:place requires "in=state:" clause to get data
#whole muni
fromJSON(paste0("https://api.census.gov/data/2010/dec/sf1?get=P001001,NAME&for=place:01010&in=state:17&key=", key_get("census-api-key")))
#partial muni
fromJSON(paste0("https://api.census.gov/data/2010/dec/sf1?get=P001001,NAME&for=county%20(or%20part):031&in=state:17%20place:03883&key=", key_get("census-api-key")))

#pull variable group from decennial
fromJSON(paste0("https://api.census.gov/data/2010/dec/sf1?get=NAME,group(P12)&for=place:01010&in=state:17&key=", key_get("census-api-key")))

#pull single estimate from ACS (rather than group as above)
fromJSON(paste0("https://api.census.gov/data/2019/acs/acs5?get=B01001_001E,NAME&for=place:01010&in=state:17&key=", key_get("census-api-key")))

#pull decennial variable for ZCTA
fromJSON(paste0("https://api.census.gov/data/2010/dec/sf1?get=P001001,NAME&for=zip%20code%20tabulation%20area:60004&key=", key_get("census-api-key")))

#pull decennial variable for part of ZCTA in Cook
fromJSON(paste0("https://api.census.gov/data/2010/dec/sf1?get=P001001,NAME&for=county%20(or%20part):031&in=state:17%20zip%20code%20tabulation%20area%20(or%20part):60466&key=", key_get("census-api-key")))

#pull decennial variable for part of ZCTA in Cook
fromJSON(paste0("https://api.census.gov/data/2010/dec/sf1?get=P001001,NAME&for=county%20(or%20part):031&in=state:17%20zip%20code%20tabulation%20area%20(or%20part):60134&key=", key_get("census-api-key")))

#try pulling for all Illinois (if error then skip, else return -- purrr version of try)

#pull all county pops from a zip code (note: can't use wildcard for zip here)
test <- fromJSON(paste0("https://api.census.gov/data/2020/dec/dhc?get=P1_001N,NAME&for=county%20(or%20part):*&in=zip%20code%20tabulation%20area:60010%20state%20(or%20part):17&key=",key_get("census-api-key")))



```


