---
title: ''
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load required packages
library(jsonlite)
library(tidyverse)
library(keyring)
library(janitor)

#store your census api key - only needs to be done once per computer
#key_set("census-api-key")

```

## Pull table B01001 (sex by age) from 5 year ACS for all CCDPH municipalities

```{r acs_b01001}

#load in ccdph muni file with place code
munis <- read_csv("https://github.com/Cook-County-Department-of-Public-Health/ccdph-data-sets/blob/main/Municipality%20Populations.csv?raw=TRUE") %>% 
  #filter(ExcludeFromAnalysis == F) %>%   
  mutate(CensusPlaceCode = ifelse(nchar(CensusPlaceCode) == 4, paste0("0", CensusPlaceCode), CensusPlaceCode))

#load in ACS variable names to match with codes
acs_vars <- fromJSON("https://api.census.gov/data/2019/acs/acs5/variables") %>%
  as.data.frame() %>%
  row_to_names(row_number = 1)

#Create empty list to store place data
acs_age_pops_muni_list <- list()

#loop through munis list to pull data
for (i in 1:nrow(munis)) {
  
  #determine if partial place string needed
  partial <- munis[[i, "Partial"]]
  
  #extract place code  -- NOTE: ifelse should not necessary if performed when importing the muni file, remove after next test
  #place_code <- as.character(munis[[i, "CensusPlaceCode"]] %>% ifelse(nchar(.) == 4, paste0("0", .), .))
  place_code <- munis[[i, "CensusPlaceCode"]]
  
  #construct api call - different string needed for if pulling for only part of town within cook county
  if (partial) {
    api <- paste0("https://api.census.gov/data/2019/acs/acs5?get=NAME,group(B01001)&for=county%20(or%20part):031&in=state:17%20place:", place_code, "&key=", key_get("census-api-key"))
  } else {
    api <- paste0("https://api.census.gov/data/2019/acs/acs5?get=NAME,group(B01001)&for=place:", place_code, "&in=state:17&key=", key_get("census-api-key"))
  }
  
  #pull data
  town_pull <- fromJSON(api) %>% 
    as.data.frame() %>%
    row_to_names(row_number = 1) %>%
    pivot_longer(cols = -NAME, names_to = "variable", values_to = "estimate") %>%
    filter(grepl("*E$", variable)) %>%    #keep only population estimates, can modify if margins of error required as well
    left_join(acs_vars, by = c("variable" = "name")) %>%   #link back to acs list of variables for label to ensure expected variables are pulled
    mutate(CensusPlaceCode = as.character(place_code))
  
  #add results to list
  acs_age_pops_muni_list[[i]] <- town_pull
  
}

#combine separate town files into one
acs_age_pops_muni <- bind_rows(acs_age_pops_muni_list)

#QUALITY CHECK -- make sure sum of pops add up to the whole
check_pops <- acs_age_pops_muni %>%
  filter(!variable %in% c("B01001_001E", "B01001_002E", "B01001_026E"))  %>% #remove total pop, total male, total female 
  group_by(NAME) %>%
  summarize(total_pop_calc = sum(as.numeric(estimate))) %>%  #sum all mutually exclusive male and female age groups for calculated total pop
  ungroup() %>%
  left_join(filter(acs_age_pops_muni, variable == "B01001_001E")) %>%
  mutate(diff = total_pop_calc - as.numeric(estimate))  #verify all diff values equal zero

#QULAITY CHECK -- make sure place code matching expected name from muni file 
name_compare <- acs_age_pops_muni %>%
  filter(variable == "B01001_001E") %>%
  left_join(munis, by = "CensusPlaceCode") %>%
  mutate(clean_census_name = gsub(" village, Illinois| city, Illinois| town, Illinois", "", NAME),
         clean_census_name = gsub("^.*, ", "", clean_census_name),
         fips_check = trimws(clean_census_name) == trimws(Municipality)) %>%
  select(NAME, CensusPlaceCode, clean_census_name, Municipality, fips_check)

#QULAITY CHECK -- compare ACS total pops to decennial total pops to check for any wild discrepancies
total_pop_compare <- acs_age_pops_muni %>%
  filter(variable == "B01001_001E") %>%
  mutate(estimate = as.numeric(estimate)) %>%
  left_join(munis, by = "CensusPlaceCode") %>%
  mutate(diff = estimate - Population2010,
         percent_change = ((estimate - Population2010) / Population2010) * 100,
         percent_change = round(percent_change, 2),
         percent_change_abs = abs(percent_change)) %>%
  select(NAME, Municipality, CensusPlaceCode, Population2010, estimate, diff, percent_change, percent_change_abs) 


#Store data set for future use -- modify storage location
#write_csv(acs_age_pops_muni, "acs-5yr-2019-age-sex-by-muni.csv")


```

## Pull table P12 (sex by age) from decennial 2010 summary file 1 for all CCDPH municipalities

*Note: API call for partial Homer Glen fails - not sure why. Call works for 2019 5 YR ACS (0 people within Cook County). Maybe decennial doesn't have any part of it in Cook County?  Filter out for now.*

```{r sf1_age_sex}

#load in ccdph muni file with place code
munis <- read_csv("https://github.com/Cook-County-Department-of-Public-Health/ccdph-data-sets/blob/main/Municipality%20Populations.csv?raw=TRUE") %>% 
  #filter(ExcludeFromAnalysis == F) %>%   
  mutate(CensusPlaceCode = ifelse(nchar(CensusPlaceCode) == 4, paste0("0", CensusPlaceCode), CensusPlaceCode)) %>%
  filter(Municipality != "Homer Glen")  #see Note above

#load in sf1 variable names to match with codes
dec_vars <- fromJSON("https://api.census.gov/data/2010/dec/sf1/variables") %>%
  as.data.frame() %>%
  row_to_names(row_number = 1)

#Create empty list to store place data
dec_age_pops_muni_list <- list()

#loop through munis list to pull data
for (i in 1:nrow(munis)) {
  
  #determine if partial place string needed
  partial <- munis[[i, "Partial"]]
  
  #extract place code  -- NOTE: ifelse should not necessary if performed when importing the muni file, remove after next test
  #place_code <- as.character(munis[[i, "CensusPlaceCode"]] %>% ifelse(nchar(.) == 4, paste0("0", .), .))
  place_code <- munis[[i, "CensusPlaceCode"]]
  
  #construct api call - different string needed for if pulling for only part of town within cook county
  if (partial) {
    api <- paste0("https://api.census.gov/data/2010/dec/sf1?get=NAME,group(P12)&for=county%20(or%20part):031&in=state:17%20place:", place_code, "&key=", key_get("census-api-key"))
  } else {
    api <- paste0("https://api.census.gov/data/2010/dec/sf1?get=NAME,group(P12)&for=place:", place_code, "&in=state:17&key=", key_get("census-api-key"))
  }
  
  #pull data
  town_pull <- fromJSON(api) %>% 
    as.data.frame() %>%
    row_to_names(row_number = 1) %>%
    pivot_longer(cols = -NAME, names_to = "variable", values_to = "estimate") %>%
    filter(grepl("^P.*", variable)) %>%    #keep only population estimates, no meta data
    filter(!grepl(".*ERR$", variable)) %>%
    left_join(dec_vars, by = c("variable" = "name")) %>%   #link back to acs list of variables for label to ensure expected variables are pulled
    mutate(CensusPlaceCode = as.character(place_code)) %>%
    mutate(estimate = as.numeric(as.character(estimate)))  #numbers being stored as factors
  
  #add results to list
  dec_age_pops_muni_list[[i]] <- town_pull
  
}

#combine separate town files into one
dec_age_pops_muni <- bind_rows(dec_age_pops_muni_list)

#QUALITY CHECK -- make sure sum of pops add up to the whole
check_pops <- dec_age_pops_muni %>%
  filter(!variable %in% c("P012001", "P012002", "P012026"))  %>% #remove total pop, total male, total female 
  group_by(NAME) %>%
  summarize(total_pop_calc = sum(estimate)) %>%  #sum all mutually exclusive male and female age groups for calculated total pop
  ungroup() %>%
  left_join(filter(dec_age_pops_muni, variable == "P012001")) %>%
  mutate(diff = total_pop_calc - as.numeric(estimate))  #verify all diff values equal zero

#QULAITY CHECK -- make sure place code matching expected name from muni file 
name_compare <- dec_age_pops_muni %>%
  filter(variable == "P012001") %>%
  left_join(munis, by = "CensusPlaceCode") %>%
  mutate(clean_census_name = gsub(" village, Illinois| city, Illinois| town, Illinois", "", NAME),
         clean_census_name = gsub("^.*, ", "", clean_census_name),
         fips_check = trimws(clean_census_name) == trimws(Municipality)) %>%
  select(NAME, CensusPlaceCode, clean_census_name, Municipality, fips_check)


#NOTE: when pulling 2020 data, may want to adapt this quality check to compare 2020 decennial to 2010 decennial for wild discrepancies
total_pop_compare <- dec_age_pops_muni %>%
  filter(variable == "P012001") %>%
  left_join(munis, by = "CensusPlaceCode") %>%
  mutate(diff = estimate - Population2010) %>%
  select(NAME, Municipality, CensusPlaceCode, Population2010, estimate, diff) 


#Store data set for future use -- modify storage location
#write_csv(dec_age_pops_muni, "decennial-2010-age-sex-by-muni.csv")

```


## Pull table P5 (Hispanic or Latino Origin by Race) from decennial 2010 summary file 1 for all CCDPH municipalities


```{r sf1_race}

#load in ccdph muni file with place code
munis <- read_csv("https://github.com/Cook-County-Department-of-Public-Health/ccdph-data-sets/blob/main/Municipality%20Populations.csv?raw=TRUE") %>% 
  #filter(ExcludeFromAnalysis == F) %>%   
  mutate(CensusPlaceCode = ifelse(nchar(CensusPlaceCode) == 4, paste0("0", CensusPlaceCode), CensusPlaceCode)) %>%
  filter(Municipality != "Homer Glen")  #see Note above age

#load in sf1 variable names to match with codes
dec_vars <- fromJSON("https://api.census.gov/data/2010/dec/sf1/variables") %>%
  as.data.frame() %>%
  row_to_names(row_number = 1)

#Create empty list to store place data
dec_race_pops_muni_list <- list()

#loop through munis list to pull data
for (i in 61:nrow(munis)) {
  
  #determine if partial place string needed
  partial <- munis[[i, "Partial"]]
  
  #extract place code  -- NOTE: ifelse should not necessary if performed when importing the muni file, remove after next test
  #place_code <- as.character(munis[[i, "CensusPlaceCode"]] %>% ifelse(nchar(.) == 4, paste0("0", .), .))
  place_code <- munis[[i, "CensusPlaceCode"]]
  
  #construct api call - different string needed for if pulling for only part of town within cook county
  if (partial) {
    api <- paste0("https://api.census.gov/data/2010/dec/sf1?get=NAME,group(P5)&for=county%20(or%20part):031&in=state:17%20place:", place_code, "&key=", key_get("census-api-key"))
  } else {
    api <- paste0("https://api.census.gov/data/2010/dec/sf1?get=NAME,group(P5)&for=place:", place_code, "&in=state:17&key=", key_get("census-api-key"))
  }
  
  #pull data
  town_pull <- fromJSON(api) %>% 
    as.data.frame() %>%
    row_to_names(row_number = 1) %>%
    pivot_longer(cols = -NAME, names_to = "variable", values_to = "estimate") %>%
    filter(grepl("^P.*", variable)) %>%    #keep only population estimates, no meta data
    filter(!grepl(".*ERR$", variable)) %>%
    left_join(dec_vars, by = c("variable" = "name")) %>%   #link back to acs list of variables for label to ensure expected variables are pulled
    mutate(CensusPlaceCode = as.character(place_code)) %>%
    mutate(estimate = as.numeric(as.character(estimate)))  #numbers being stored as factors
  
  #add results to list
  dec_race_pops_muni_list[[i]] <- town_pull
  
}

#combine separate town files into one
dec_race_pops_muni <- bind_rows(dec_race_pops_muni_list)

#QUALITY CHECK -- make sure sum of pops add up to the whole
check_pops <- dec_race_pops_muni %>%
  filter(!variable %in% c("P005001", "P005002", "P005010"))  %>% #remove total pop, total non-hispanic, total hispanic 
  group_by(NAME) %>%
  summarize(total_pop_calc = sum(estimate)) %>%  #sum all mutually exclusive race/ethnicity groups for calculated total pop
  ungroup() %>%
  left_join(filter(dec_race_pops_muni, variable == "P005001")) %>%
  mutate(diff = total_pop_calc - estimate)  #verify all diff values equal zero

#QULAITY CHECK -- make sure place code matching expected name from muni file 
name_compare <- dec_race_pops_muni %>%
  filter(variable == "P005001") %>%
  left_join(munis, by = "CensusPlaceCode") %>%
  mutate(clean_census_name = gsub(" village, Illinois| city, Illinois| town, Illinois", "", NAME),
         clean_census_name = gsub("^.*, ", "", clean_census_name),
         fips_check = trimws(clean_census_name) == trimws(Municipality)) %>%
  select(NAME, CensusPlaceCode, clean_census_name, Municipality, fips_check)


#QULAITY CHECK -- compare ACS total pops to decennial total pops to check for any wild discrepancies
#NOTE: when pulling 2020 data, may want to adapt this quality check to compare 2020 decennial to 2010 decennial for wild discrepancies
total_pop_compare <- dec_race_pops_muni %>%
  filter(variable == "P00501") %>%
  left_join(munis, by = "CensusPlaceCode") %>%
  mutate(diff = estimate - Population2010) %>%
  select(NAME, Municipality, CensusPlaceCode, Population2010, estimate, diff) 

#Store data set for future use -- modify storage location
#write_csv(dec_race_pops_muni, "decennial-2010-race-by-muni.csv")


```



```{r general_examples}

#pull decennial variable for all of cook county
fromJSON(paste0("https://api.census.gov/data/2010/dec/sf1?get=P001001,NAME&for=county:031&in=state:17&key=", key_get("census-api-key")))

#pull decennial variable for place within cook county -- NOTE:place requires "in=state:" clause to get data
#whole muni
fromJSON(paste0("https://api.census.gov/data/2010/dec/sf1?get=P001001,NAME&for=place:01010&in=state:17&key=", key_get("census-api-key")))
#partial muni
fromJSON(paste0("https://api.census.gov/data/2010/dec/sf1?get=P001001,NAME&for=county%20(or%20part):031&in=state:17%20place:03883&key=", key_get("census-api-key")))

#pull variable group from decennial
fromJSON(paste0("https://api.census.gov/data/2010/dec/sf1?get=NAME,group(P12)&for=place:01010&in=state:17&key=", key_get("census-api-key")))

#pull single estimate from ACS (rather than group as above)
fromJSON(paste0("https://api.census.gov/data/2019/acs/acs5?get=B01001_001E,NAME&for=place:01010&in=state:17&key=", key_get("census-api-key")))

#pull decennial variable for ZCTA
fromJSON(paste0("https://api.census.gov/data/2010/dec/sf1?get=P001001,NAME&for=zip%20code%20tabulation%20area:60004&key=", key_get("census-api-key")))

#pull decennial variable for part of ZCTA in Cook
fromJSON(paste0("https://api.census.gov/data/2010/dec/sf1?get=P001001,NAME&for=county%20(or%20part):031&in=state:17%20zip%20code%20tabulation%20area%20(or%20part):60466&key=", key_get("census-api-key")))

#pull decennial variable for part of ZCTA in Cook
fromJSON(paste0("https://api.census.gov/data/2010/dec/sf1?get=P001001,NAME&for=county%20(or%20part):031&in=state:17%20zip%20code%20tabulation%20area%20(or%20part):*&key=", key_get("census-api-key")))



```


