---
title: ''
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load required packages
library(jsonlite)
library(tidyverse)
library(keyring)
library(janitor)

#store your census api key - only needs to be done once per computer
#key_set("census-api-key")

```

## Import current zip code file - "notes" field contains institutional knowledge / convention and must be retained and integrated into new file

```{r current_zips}

current_zips <- read_csv("cook-county-zip-codes.csv") %>%
  select(zip_code, 
         notes)

```

## Pull list of all Cook zips from Melissa data

Note: Several years of working with Melissa data have not revealed any inaccuracies with the USPS data but USPS does have city to zip code and zip code to city look-ups that could be used to populate this information directly from USPS in the future using RSelenium code

```{r pull_zip_list}

#Note: error returned when pulling with jsonlite so using rjsonio instead
melissa_cook_zips <- RJSONIO::fromJSON("https://www.melissa.com/v2/lookups/countyzip/state/?fips=17031&fmt=json") %>%
  map(as.data.frame) %>% 
  bind_rows() %>%
  clean_names() %>%
  mutate(primary_city = str_to_title(city),
         zip = as.numeric(as.character(zip))) %>%
  select(zip,
         primary_city,
         percent_addresses = percentage,  #per Melissa data, percentage reflects percent addresses in cook county
         type)

```

## Pull total population from decennial 2010 summary file 1 for all ZCTAs that have population in Cook County

```{r sf1_total_zcta}

#list zctas in illinois (note:only one IL ZCTA crosses state lines, down state so not of concern for CCDPH)
zcta_il <- fromJSON(paste0("https://api.census.gov/data/2010/dec/sf1?get=P001001,NAME&for=zip%20code%20tabulation%20area%20(or%20part):*&in=state:17&key=", key_get("census-api-key"))) %>%
  as.data.frame() %>%
  row_to_names(row_number = 1) %>%
  select(zcta = `zip code tabulation area (or part)`,
         population = P001001)

#function to pull pop in cook
get_partial <- function(zcta){
  
  value <- fromJSON(paste0("https://api.census.gov/data/2010/dec/sf1?get=P001001,NAME&for=county%20(or%20part):031&in=state:17%20zip%20code%20tabulation%20area%20(or%20part):", zcta, "&key=", key_get("census-api-key")))[2,1]
  
  return(as.numeric(value))
  
}

#Note: code below works but is very slow (~25 minutes); might want to for loop with try() at some point
zcta_pops <- zcta_il %>%
  rowwise() %>%
  mutate(population_cook = possibly(get_partial, otherwise = NA)(zcta))

zcta_pops_clean <- zcta_pops %>%
  filter(!is.na(population_cook) | zcta %in% melissa_cook_zips$zip) %>%
  mutate(across(c(zcta, population), ~as.numeric(as.character(.x)))) %>%
  mutate(percent_pop = (population_cook / population) * 100,
         percent_pop = round(percent_pop, 2),
         partial_pop = population != population_cook) %>%
  full_join(melissa_cook_zips, by = c("zcta" = "zip")) %>%
  rename(zcta_2010_pop = population, zcta_2010_pop_cook = population_cook)

```

## Import NSSP region to zip code mappings

```{r nssp}

#add nssp zip code region to file (most recent copy received June 2021)
nssp <- readxl::read_excel(paste0(key_get("nssp-folder"), "IL_Zipcode_Region_Mappings.xlsx")) %>% 
  clean_names() %>% 
  select(-lat, -lon) %>%
  filter(region %in% c("IL_Cook", "IL_Chicago") | zip_code %in% zcta_pops_clean$zcta)

```

## Combine all necessary pieces of information into final file

```{r final}

#import municipality file - zip code district will be determined by the district of the primary city
districts <- read_csv("https://github.com/Cook-County-Department-of-Public-Health/ccdph-data-sets/blob/main/Municipality%20Populations.csv?raw=TRUE") %>%
  select(Municipality, District) %>%
  janitor::clean_names()

final_zips <- zcta_pops_clean %>%
  full_join(current_zips, by = c("zcta" = "zip_code")) %>%
  mutate(notes = case_when(is.na(notes) & type == "PO Boxes" ~ "PO Boxes",
                           !is.na(notes) & type == "PO Boxes" ~ paste0(notes, "; PO Boxes"),
                           TRUE ~ notes)) %>%
  select(-type) %>%
  full_join(nssp, by = c("zcta" = "zip_code")) %>%
  mutate(primary_city = case_when(!is.na(primary_city) ~ primary_city,
                                  grepl("^606.*", zcta) ~ "Chicago",
                                  grepl("^603.*", zcta) ~ "Oak Park",
                                  grepl("^602.*", zcta) ~ "Evanston")) %>%
  left_join(districts, by = c("primary_city" = "municipality")) %>%
  mutate(district = case_when(!is.na(district) ~ district,
                              primary_city == "Summit Argo" ~ "Southwest",
                              primary_city == "Techny" ~ "North",
                              primary_city == "Hines" ~ "West",
                              !grepl("Cook|Chicago", region) ~ "OOJ")) %>%
  select(zip_code = zcta,
         zcta_2010_pop:percent_addresses,
         district,
         nssp_region = region,
         notes) %>%
  mutate(zcta = ifelse(!is.na(zcta_2010_pop), zip_code, NA)) %>%
  relocate(zcta, .after = zip_code)


#write csv and fill in missing information manually
write_csv(final_zips, "cook-county-zip-codes.csv")
  

```


## Cross reference against GIS Department list of zip codes (purchased from vendor in early 2021)

```{r gis_check}

#Import attribute table exported from GIS Department zip code map
gis_zips <- readxl::read_excel(paste0(key_get("onedrive-path"),"COVID Home/Home working temp/gis_zips.xlsx"))

#Compare against final list
gis_missing <- bot_zips[!bot_zips$ZIP %in% final_zips$zip_code, "ZIP"] 
#missing zips are primaily out of jurisdiction - no people in Cook for corresponding zip codes

#Add fields for concatenating with current list
gis_missing_clean <- gis_missing %>%
  rename(zip_code = ZIP) %>%
  mutate(zcta = zip_code) %>%
  left_join(zcta_il) %>% 
  mutate(zcta_2010_pop = as.numeric(population)) %>%
  select(-population) %>%
  mutate(zip_code = as.numeric(zip_code),
         zcta = as.numeric(zcta),
         zcta_2010_pop_cook = 0,
         percent_pop = 0,
         partial_pop = NA,
         primary_city = NA,
         percent_addresses = 0,
         district = "OOJ") %>%
  left_join(nssp) %>%  #re-imported excel from above with no filter
  rename(nssp_region = region) %>%
  mutate(notes = "from BOT list but no addresses in Cook per melissa data")
  
#add data to final file for GIS zips - primary city will be filled in manually from USPS lookups
final_zips <- read_csv("cook-county-zip-codes.csv") %>%
  bind_rows(gis_missing_clean)
write_csv(final_zips, "cook-county-zip-codes.csv")

```



## Manual updates post processing

1) Notes were added for non-PO box zip codes with no corresponding ZCTA based on USPS look ups and Google searches. 

